---
title: "An update to the global critical habitat screening layer"
author:
  - name: Sebastian Dunnett
    orcid: 0000-0002-4238-2508
    email: sebastian.dunnett@unep-wcmc.org
    affiliations:
      - name: United Nations Environment Programme World Conservation Monitoring Centre
        address: 219 Huntingdon Road
        city: Cambridge
        postal-code: CB3 0DL
        id: unep-wcmc
  - name: Alfred Muge
    affiliations:
      - ref: unep-wcmc
  - name: Alex Ross
    affiliations:
      - ref: unep-wcmc
  - name: Joe Turner
    affiliations:
      - name: The Biodiversity Consultancy
        address: 3E King's Parade
        city: Cambridge
        postal-code: CB2 1SJ
        id: tbc
  - name: Neil Burgess
    affiliations:
      - ref: unep-wcmc
  - name: Matt Jones
    affiliations:
      - ref: unep-wcmc
format: html
bibliography: references.bib
abstract: "Critical Habitat has become an increasingly important concept used by the finance sector and businesses to identify areas of high biodiversity value. The International Finance Corporation (IFC) defines Critical Habitat in their highly influential Performance Standard 6 (PS6), requiring projects in Critical Habitat to achieve a net gain of biodiversity. Here we present a global screening layer of Critical Habitat in the terrestrial realm, derived from global spatial datasets covering the distributions of 12 biodiversity features aligned with guidance provided by the IFC. Each biodiversity feature is categorised as ‘likely’ or ‘potential’ Critical Habitat based on: 1. Alignment between the biodiversity feature and the IFC Critical Habitat definition; and 2. Suitability of the spatial resolution for indicating a feature’s presence on the ground. Following the initial screening process, Critical Habitat must then be assessed in-situ by a qualified assessor. This analysis indicates that a total of 10% and 5% of the global terrestrial environment can be considered as likely and potential Critical Habitat, respectively, while the remaining 85% did not overlap with any of the biodiversity features assessed and was classified as ‘unknown’. Likely Critical Habitat was determined principally by the occurrence of Key Biodiversity Areas and Protected Areas. Potential Critical Habitat was predominantly characterised by data representing highly threatened and unique ecosystems such as ever-wet tropical forests and tropical dry forests. The areas we identified as likely or potential Critical Habitat are based on the best available global-scale data for the terrestrial realm that is aligned with IFC’s Critical Habitat definition. Our results can help businesses screen potential development sites at the early project stage based on a range of biodiversity features. However, the study also demonstrates several important data gaps and highlights the need to incorporate new and improved global spatial datasets as they become available. Test = `r 1+1`"
---

```{r setup}
#| include: false

# Install packages (if required)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,terra,sf,giscoR,scico,cowplot,kableExtra,scales)
terraOptions(progress=0,memfrac=0.9)

ch_pal = c("#8400a8","#ff00c5","#e8e8e8")

```

```{r import-rasters}
#| include: false

raster_WGS <- rast(res=1/12)

raster_eq <- project(raster_WGS,"EPSG:8857",res=10000)

sphere <- st_graticule(ndiscr = 10000, margin = 10e-6) %>%
  st_transform(crs = st_crs("ESRI:54009")) %>%
  st_convex_hull() %>%
  summarise(geometry = st_union(geometry)) |>
  st_buffer(1000) |> st_union() |> st_buffer(-1000) |>
  st_transform("EPSG:8857")

ch2023_1km = rast("../../outputs/Basic_Critical_Habitat_Raster_WGS.tif")

ch2023_areas = expanse(ch2023_1km,unit="km",byValue=TRUE)[,2:3]

ch2023 = rast("../../outputs/paper/ch2023_10km.tif") |>
project(raster_eq,method="near",mask=TRUE)
levels(ch2023) = data.frame(id=c(0,1,10), ch2023=c("Unclassified","Potential Critical Habitat","Likely Critical Habitat"))

ch2018 = rast("../../outputs/paper/ch2018_10km.tif") |>
project(raster_eq,method="near",mask=TRUE)
levels(ch2018) = data.frame(id=c(0,1,10), ch2018=c("Unclassified","Potential Critical Habitat","Likely Critical Habitat"))

ch_combined = c(ch2018,ch2023)
names(ch_combined) = c("2018","2023")

ch_combined_df = as.data.frame(ch_combined,xy=TRUE) |>
  pivot_longer(3:4,names_to="year",values_to="ch") |>
  mutate(year=factor(year,levels=c("2018","2023")),
         ch=factor(ch,levels=c("Likely Critical Habitat","Potential Critical Habitat","Unclassified")))
  
ch_changes = rast("../../outputs/paper/changes_2018_2023.tif") |>
  aggregate(fact=10,fun="modal") |>
  project(raster_eq,method="near",mask=TRUE)

ch_changes_df = as.data.frame(ch_changes,xy=TRUE) |>
  rename("change"=3) |>
  mutate(change=case_match(change,
                           -2 ~ "Removal",
                           -1 ~ "Downgrade",
                           0 ~ "No change",
                           1 ~ "Upgrade",
                           2 ~ "Addition") |>
           factor(levels=c("Removal","Downgrade","No change","Upgrade","Addition")))

wrld = st_read("../../raw_data/UN map borders/Countries_Combined_with_associated_territories.json",quiet=TRUE) |>
  st_wrap_dateline() |>
  st_transform("EPSG:8857") |>
  st_simplify(dTolerance=1000)

```

## Background & Summary

The **International Finance Corporation (IFC)**, a member of the World Bank Group, recognizes the importance of protecting critical habitat as part of its commitment to sustainable development and environmental stewardship. Critical habitat refers to the areas that are essential for the survival and well-being of species and ecosystems [@brauneder2018].

1.  Critically Endangered or Endangered species
2.  Endemic and/or restricted-range species
3.  Globally significant concentration of migratory or congregatory species
4.  Highly threatened and/or unique ecosystems
5.  Key evolutionary processes

## Methods

```{r class-grid}
#| fig-width: 8
#| fig-height: 8
#| echo: false

classGrid = data.frame(xmin=c(0,0,0.5,0.5),
                xmax=c(0.5,0.5,1,1),
                ymin=c(0,0.5,0,0.5),
                ymax=c(0.5,1,0.5,1),
                ch=c("Unclassified","Potential Critical Habitat","Potential Critical Habitat","Likely Critical Habitat"),
                labx=c(0.25,0.25,0.75,0.75),
                laby=c(0.25,0.75,0.25,0.75))

c = ggplot(classGrid) +
  geom_rect(aes(xmin=xmin,xmax=xmax,ymin=ymin,ymax=ymax,fill=ch)) +
  geom_text(aes(label=ch,x=labx,y=laby),fontface="bold") +
  theme_void() +
  scale_fill_manual(values=ch_pal) +
  theme(legend.position = "none",
        plot.margin = unit(c(rep(0.5,4)),"in"),
        axis.line.y = element_line(arrow=grid::arrow(ends="last",type="open",length=unit(0.1,"in"))),
        axis.line.x = element_line(arrow=grid::arrow(ends="last",type="open",length=unit(0.1,"in"))),
        axis.title.x = element_text(),
        axis.title.y = element_text(angle=90)) +
  xlab("\nAlignment / relevance to IFC PS6") +
  ylab("Presence on the ground\n")

ggdraw(c) +
  draw_text("Potential",x=0.15,y=0.07,size=13,fontface="italic") +
  draw_text("Likely",x=0.9,y=0.07,size=13,fontface="italic") +
  draw_text("Uncertain",x=0.075,y=0.15,size=13,fontface="italic",angle=90) +
  draw_text("Certain",x=0.075,y=0.9,size=13,fontface="italic",angle=90)

```

## Data Records

## Technical validation

## Usage notes

```{r main-plot}
#| fig-width: 8
#| fig-height: 7
#| echo: false

a = ggplot(ch_combined_df) +
  geom_raster(aes(x=x,y=y,fill=ch)) +
  geom_sf(data=st_transform(gisco_coastallines,"EPSG:8857"),fill=NA) +
  theme_void() +
  facet_wrap(~year) +
  scale_fill_manual(values=ch_pal,name=NULL) + 
  theme(legend.position="bottom")

b = ggplot(ch_changes_df) +
  geom_raster(aes(x=x,y=y,fill=factor(change))) +
  geom_sf(data=st_transform(gisco_coastallines,"EPSG:8857"),fill=NA) +
  theme_void() +
  scale_fill_scico_d(palette="bam",name=NULL) +
  theme(legend.position="bottom")

plot_grid(a,b,rel_heights=c(0.3,0.7),nrow=2)

```

```{r ch-coverage-table}
#| echo: false
#| tbl-cap: "Areas and percentage coverage of each Critical Habitat category in 2018 and 2023 across Areas Beyond National Jurisdiction (ABNJ), Economic Exclusion Zones (EEZ) and Land."

wrld_ch_areas = read.csv("wrld_ch_areas.csv",check.names = F) |>
  group_by(type) |>
  summarise(across(starts_with(c("2018","2023")),~round(sum(.x,na.rm=TRUE)/1000,digits=2)))

wrld_ch_pct = wrld_ch_areas |>
  reframe(across(2:7,~label_percent(accuracy=.1)(.x/sum(.x)))) |>
  rename_with(~paste0(.x,"_pct"))

wrld_ch_areas_pct = cbind(wrld_ch_areas,wrld_ch_pct) %>% 
  select(type,sort(names(.)))

kable(wrld_ch_areas_pct,
      col.names = c("Domain","Likely","%","Potential","%","Unclassified","%","Likely","%","Potential","%","Unclassified","%")) |>
  add_header_above(header=c(" "=1,"2018"=6,"2023"=6))

```

```{r criteria-areas-table}
#| echo: false
#| tbl-cap: "Areas and percentage cover of each IFC PS6 criteria to Potential and Likely Critical Habitat. Percentages add up to >100 due to overlapping areas."

criteria_areas = read.csv("drill_down_change_areas.csv", check.names = F) |>
  select(CH,C1:C5,area) |>
  pivot_longer(C1:C5,values_to="present",names_to="criteria") |>
  filter(present>0) |>
  group_by(criteria,CH) |>
  summarise(area=sum(area)/1000) |>
  pivot_wider(id_cols=1,names_from="CH",values_from="area") |>
  ungroup() |>
  mutate(total_potential=ch2023_areas[ch2023_areas$value==1,2],
         total_likely=ch2023_areas[ch2023_areas$value==10,2],
         likely_pct=label_percent()(`10`/(total_likely/1000)),
         potential_pct=label_percent()(`1`/(total_potential/1000)),
         criteria=case_match(criteria,
                             "C1" ~ "1. Critically Endangered or Endangered species",
                             "C2" ~ "2. Endemic and/or restricted-range species",
                             "C3" ~ "3. Globally significant concentration of migratory or congregatory species",
                             "C4" ~ "4. Highly threatened and/or unique ecosystems",
                             "C5" ~ "5. Key evolutionary processes")) |>
  select(criteria,`10`,likely_pct,`1`,potential_pct)

kable(criteria_areas,
      col.names = c("Criteria","","","",""),
      format.args = list(digits=4,big.mark=",")) |>
  add_header_above(c(" "=1,"Likely"=2,"Potential"=2))

```

```{r features-table}
#| echo: false
#| tbl-cap: "Biodiversity features included in the analysis, their alignment with IFC PS6 Critical Habitat criteria and classification as 'Likely' or 'Potential' Critical Habitat. 'CR': Critically Endangered; 'EN': Endangered; 'L': Likely; 'P': Potential."

fts = read.csv("features_table.csv") |>
  rename(`Biodiversity features`=1,
         `Designation criterion / Trigger`=2,
         `1`=3,`2`=4,`3`=5,`4`=6,`5`=7,
         `Classification`=8)

fts |>
  kable() |>
  add_header_above(header=c(" "=2,"IFC PS6 criteria"=5," "=1)) |>
  column_spec(3, color = ifelse(fts$`1` == "L", "darkorchid2", ifelse(fts$`1` == "P","magenta1",FALSE)), bold = TRUE) |>
  column_spec(4, color = ifelse(fts$`2` == "L", "darkorchid2", ifelse(fts$`2` == "P","magenta1",FALSE)), bold = TRUE) |>
  column_spec(5, color = ifelse(fts$`3` == "L", "darkorchid2", ifelse(fts$`3` == "P","magenta1",FALSE)), bold = TRUE) |>
  column_spec(6, color = ifelse(fts$`4` == "L", "darkorchid2", ifelse(fts$`4` == "P","magenta1",FALSE)), bold = TRUE) |>
  column_spec(7, color = ifelse(fts$`5` == "L", "darkorchid2", ifelse(fts$`5` == "P","magenta1",FALSE)), bold = TRUE) |>
  collapse_rows(1)
  

```
