---
title: "An update to the global Critical Habitat screening layer"
author:
  - name: Sebastian Dunnett
    orcid: 0000-0002-4238-2508
    email: sebastian.dunnett@unep-wcmc.org
    affiliations:
      - name: United Nations Environment Programme World Conservation Monitoring Centre
        address: 219 Huntingdon Road
        city: Cambridge
        country: UK
        postal-code: CB3 0DL
        id: unep-wcmc
  - name: Alfred Muge
    affiliations:
      - ref: unep-wcmc
  - name: Alex Ross
    affiliations:
      - ref: unep-wcmc
  - name: Joseph A. Turner
    affiliations:
      - name: The Biodiversity Consultancy
        address: 3E King's Parade
        city: Cambridge
        country: UK
        postal-code: CB2 1SJ
        id: tbc
  - name: Neil D. Burgess
    orcid: 0000-0003-2291-4535
    affiliations:
      - ref: unep-wcmc
  - name: Matt Jones
    orcid: 0000-0002-5690-4274
    affiliations:
      - ref: unep-wcmc
  - name: Sharon Brooks
    orcid: 0000-0001-8744-3557
    affiliations:
      - ref: unep-wcmc
format: scidata-pdf
bibliography: references.bib
abstract: "The International Finance Corporation (IFC) defines Critical Habitat in Performance Standard 6 (PS6) as high biodiversity value areas requiring net biodiversity gain for projects. We present an updated global screening layer of Critical Habitat aligned with IFC's 2019 guidance. This layer derives from global datasets covering 54 biodiversity features, categorized as 'Likely' or 'Potential' Critical Habitat based on agreement with IFC criteria and data suitability. Analysis indicates 53.95 million km^2^ (10.58%) and 13.71 million km^2^ (2.69%) of the globe can be considered Likely and Potential Critical Habitat respectively, with the remaining 86.73% not overlapping with assessed biodiversity features. This represents a significant increase over previously identified 5% and 2% for Likely and Potential Critical Habitat. Likely Critical Habitat was dominated by Important Bird and Biodiversity Areas, Intact Forest Landscapes, and protected areas; Potential Critical Habitat by Important Marine Mammal Areas and ranges of IUCN Vulnerable species. Our results can help businesses prioritize impact avoidance and identify opportunities by screening potential development sites for biodiversity features."
---

```{r setup}
#| cache: false

# Install packages (if required)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,terra,sf,giscoR,scico,cowplot,kableExtra,scales,units,tidyterra,httr,jsonlite)
terraOptions(progress=0)

ch_pal = c("#8400a8","#ff00c5","#e8e8e8")

knitr::knit_hooks$set(inline = function(x) {
  if(!is.numeric(x)){x}
  else{
    prettyNum(signif(x,4), big.mark=",")
    }
  })

```

```{r import-rasters}
#| cache: false

raster_WGS <- rast(res=1/12)

raster_eq <- project(raster_WGS,"EPSG:8857",res=10000)

sphere <- st_graticule(ndiscr = 10000, margin = 10e-6) %>%
  st_transform(crs = st_crs("ESRI:54009")) %>%
  st_convex_hull() %>%
  summarise(geometry = st_union(geometry)) |>
  st_buffer(1000) |> st_union() |> st_buffer(-1000) |>
  st_transform("EPSG:8857")

ch_basic_files = list.files("..", pattern = "Basic_Critical_Habitat.*\\.tif$", full.names=TRUE)
ch_basic_file = ch_basic_files[which.max(file.info(ch_basic_files)$ctime)]

ch_poly_files = list.files("..", pattern = "Drill_Down_Critical_Habitat_Polygons.*\\.gpkg$", full.names=TRUE)
ch_poly_file = ch_poly_files[which.max(file.info(ch_poly_files)$ctime)]

ch2025_30as = rast(ch_basic_file)

ch2025_areas = expanse(ch2025_30as,unit="km",byValue=TRUE)[,2:3]
ch2018_areas = expanse(rast("ch2018_30as.tif"),unit="km",byValue=TRUE)[,2:3]

ch2025 = rast("../../outputs/paper/ch2025_1am.tif") |>
project(raster_eq,method="near",mask=TRUE)
levels(ch2025) = data.frame(id=c(0,1,10), ch2025=c("Unclassified","Potential Critical Habitat","Likely Critical Habitat"))

ch2018 = rast("../../outputs/paper/ch2018_1am.tif") |>
project(raster_eq,method="near",mask=TRUE)
levels(ch2018) = data.frame(id=c(0,1,10), ch2018=c("Unclassified","Potential Critical Habitat","Likely Critical Habitat"))

ch_combined = c(ch2018,ch2025)
names(ch_combined) = c("2018","2025")

ch_combined_df = as.data.frame(ch_combined,xy=TRUE) |>
  pivot_longer(3:4,names_to="year",values_to="ch") |>
  mutate(year=factor(year,levels=c("2018","2025")),
         ch=factor(ch,levels=c("Likely Critical Habitat","Potential Critical Habitat","Unclassified")))
  
ch_changes = rast("../../outputs/paper/changes_2018_2025.tif") |>
  aggregate(fact=10,fun="modal") |>
  project(raster_eq,method="near",mask=TRUE)

ch_changes_areas = read.csv("wrld_ch_areas.csv",check.names = F) |>
  select(starts_with("ch_")) |>
  pivot_longer(everything(),values_to="ch_area",names_to="ch_type") |>
  group_by(ch_type) |>
  summarise(total_area=sum(ch_area,na.rm=TRUE)) |>
  ungroup()

ch_changes_df = as.data.frame(ch_changes,xy=TRUE) |>
  rename("change"=3) |>
  mutate(change=case_match(change,
                           -2 ~ "Removed",
                           -1 ~ "Downgraded",
                           0 ~ "No change",
                           1 ~ "Upgraded",
                           2 ~ "Additional") |>
           factor(levels=c("Removed","Downgraded","No change","Upgraded","Additional")))

wrld = gisco_get_coastallines(resolution="03") |>
  st_transform("EPSG:8857") |>
  st_simplify(dTolerance=1000)

features_table = read.csv("features_table.csv",check.names=F)

```

```{r class-grid}
#| fig-width: 5
#| fig-height: 5
#| label: fig-class-grid
#| fig-cap: "Classification of data as Likely or Potential Critical Habitat is based on the perceived strength of alignment with PS6 criteria and the spatial resolution of the data. Adapted from Brauneder et al. [@braunederGlobalScreeningCritical2018]."
#| echo: false

classGrid = data.frame(xmin=c(0,0,0.5,0.5),
                xmax=c(0.5,0.5,1,1),
                ymin=c(0,0.5,0,0.5),
                ymax=c(0.5,1,0.5,1),
                ch=c("Unclassified","Potential Critical Habitat","Potential Critical Habitat","Likely Critical Habitat"),
                labx=c(0.25,0.25,0.75,0.75),
                laby=c(0.25,0.75,0.25,0.75))

c = ggplot(classGrid) +
  geom_rect(aes(xmin=xmin,xmax=xmax,ymin=ymin,ymax=ymax,fill=ch)) +
  geom_label(aes(label=ch,x=labx,y=laby),fontface="bold",size=9,size.unit="pt",fill=ch_pal[3]) +
  theme_void() +
  scale_fill_manual(values=ch_pal) +
  theme(legend.position = "none",
        plot.margin = unit(c(rep(0.5,4)),"in"),
        axis.line.y = element_line(arrow=grid::arrow(ends="last",type="open",length=unit(0.1,"in"))),
        axis.line.x = element_line(arrow=grid::arrow(ends="last",type="open",length=unit(0.1,"in"))),
        axis.title.x = element_text(),
        axis.title.y = element_text(angle=90)) +
  xlab("\nAlignment / relevance to IFC PS6") +
  ylab("Presence on the ground\n")

ggdraw(c) +
  draw_text("Potential",x=0.15,y=0.07,size=11,fontface="italic") +
  draw_text("Likely",x=0.9,y=0.07,size=11,fontface="italic") +
  draw_text("Uncertain",x=0.075,y=0.15,size=11,fontface="italic",angle=90) +
  draw_text("Certain",x=0.075,y=0.9,size=11,fontface="italic",angle=90)

```

We identified `r length(unique(features_table[,1]))` biodiversity feature datasets that are split into `r length(unique(paste0(features_table[,1],features_table[,2])))` separate triggers.

We find that `r sum(ch2025_areas[2:3,2])/1000000` million km^2^ of the Earth's surface was classified as Likely or Potential Critical Habitat: `r ch2025_areas[3,2]/1000000` million km^2^ (`r label_percent(accuracy=.01)(ch2025_areas[3,2]/sum(ch2025_areas[,2]))`) as Likely Critical Habitat and `r ch2025_areas[2,2]/1000000` million km^2^ (`r label_percent(accuracy=.01)(ch2025_areas[2,2]/sum(ch2025_areas[,2]))`) as Potential Critical Habitat. This is a significant increase on the `r ch2018_areas[3,2]/1000000` million km^2^ (`r label_percent(accuracy=.01)(ch2018_areas[3,2]/sum(ch2018_areas[,2]))`) and `r ch2018_areas[2,2]/1000000` million km^2^ (`r label_percent(accuracy=.01)(ch2018_areas[2,2]/sum(ch2018_areas[,2]))`) previously identified. The remaining `r ch2025_areas[1,2]/1000000` million km^2^ (`r label_percent(accuracy=.01)(ch2025_areas[1,2]/sum(ch2025_areas[,2]))`) is "Unclassified" as either known biodiversity features do not align with the IFC definition or because appropriate data that might be used to classify do not exist: Critical Habitat may still occur in these regions.

```{r main-plot}
#| fig-width: 8
#| fig-height: 7
#| fig-cap: Global screening layer for Critical Habitat. Reprojected to Equal Earth and aggregated to 10 x 10 km for visualisation.
#| label: fig-main
#| fig-pos: 'H'

a = ggplot(ch_combined_df |> drop_na(ch)) +
  geom_raster(aes(x=x,y=y,fill=ch)) +
  geom_sf(data=st_transform(gisco_coastallines,"EPSG:8857"),fill=NA) +
  theme_void() +
  facet_wrap(~year) +
  scale_fill_manual(values=ch_pal,name=NULL) + 
  theme(legend.position="bottom")

b = ggplot(ch_changes_df) +
  geom_raster(aes(x=x,y=y,fill=factor(change))) +
  geom_sf(data=st_transform(gisco_coastallines,"EPSG:8857"),fill=NA) +
  theme_void() +
  scale_fill_scico_d(palette="bam",name=NULL) +
  theme(legend.position="bottom")

plot_grid(a,b,rel_heights=c(0.3,0.7),nrow=2)

```

```{r region-plot}
#| fig-width: 8
#| fig-height: 7
#| fig-cap: Regional trends in Critical Habitat changes.
#| label: fig-region

ipbes_wrld = st_read("../../raw_data/EEZv8_WVS_DIS_V3_ALL_final_v7disIPBES",quiet=TRUE)
perCountry = read.csv("regional_perCountry.csv")

sf_use_s2(FALSE)
plot_wrld = st_crop(ipbes_wrld,st_bbox(gisco_get_countries()))
sf_use_s2(TRUE)

plot_wrld = st_transform(plot_wrld,"EPSG:8857") |>
  st_simplify(dTolerance=1000)

wrld_perCountry = cbind(plot_wrld,perCountry[,-1])

a = ggplot(plot_wrld) +
  geom_sf(aes(fill=IPBES_regi,alpha=type),col=NA) +
  geom_sf(data=filter(plot_wrld,type=="Land"),fill=NA) +
  scale_fill_manual(name=NULL,values=c("lightblue","#D85C71","#E0E15A","#7CDF9A","#AC41E2","gray")) +
  scale_alpha_manual(values=c(.6,.6,1),guide="none") +
  cowplot::theme_map() +
  theme(legend.position = "bottom",
        legend.key.size = unit(0.5, 'cm'),
        legend.key.spacing.x = unit(0.1,"lines"),
        legend.key.spacing.y = unit(0.1,"lines"),
        legend.title.position = "top",
        legend.title = element_text(size=7,hjust=0.5),
        legend.text = element_text(size=6, margin=margin(l=0,r=0)),
        legend.justification.bottom = "center",
        legend.box.spacing = unit(0,"lines"))

b = ggplot(wrld_perCountry, aes(fill=removed)) +
  geom_sf(col=NA) +
  scale_fill_scico(name=NULL, palette="acton", labels = scales::label_percent(accuracy=1)) +
  theme_void() +
  guides(fill = guide_colourbar(theme = theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.text.position = "left"
    )))

c = ggplot(wrld_perCountry, aes(fill=downgraded)) +
  geom_sf(col=NA) +
  scale_fill_scico(name=NULL, palette="acton", labels = scales::label_percent(accuracy=.1)) +
  theme_void() +
  guides(fill = guide_colourbar(theme = theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.text.position = "left"
  )))

d = ggplot(wrld_perCountry, aes(fill=nochange)) +
  geom_sf(col=NA) +
  scale_fill_scico(name=NULL, palette="acton", labels = scales::label_percent(accuracy=1)) +
  theme_void() +
  guides(fill = guide_colourbar(theme = theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.text.position = "left"
  )))

e = ggplot(wrld_perCountry, aes(fill=upgraded)) +
  geom_sf(col=NA) +
  scale_fill_scico(name=NULL, palette="acton", labels = scales::label_percent(accuracy=1)) +
  theme_void() +
  guides(fill = guide_colourbar(theme = theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.text.position = "left"
  )))

f = ggplot(wrld_perCountry, aes(fill=added)) +
  geom_sf(col=NA) +
  scale_fill_scico(name=NULL, palette="acton", labels = scales::label_percent(accuracy=1)) +
  theme_void() +
  guides(fill = guide_colourbar(theme = theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.text.position = "left"
  )))

plot_grid(a,b,c,d,e,f,ncol=2,labels=letters[1:6])

```

There are `r nrow(st_read(ch_poly_file,quiet=TRUE))` polygons in the GeoPackage.

```{r realm-pcts}

realm_pcts = read.csv("wrld_ch_areas.csv",check.names = F) |>
  group_by(type) |>
  summarise(across(c(starts_with(c("2018","2025")),total_country_area_km2),~sum(.x,na.rm=TRUE)/1000000)) |>
  mutate(`2025_Total` = `2025_Likely` + `2025_Potential`, `2018_Total` = `2018_Likely` + `2018_Potential`) |>
  mutate(across(starts_with(c("2018","2025")), ~ .x / total_country_area_km2)) |>
  dplyr::select(c(type,ends_with("Total"))) |>
  mutate(Increase=`2025_Total`/`2018_Total`)

```

Both categories of Critical Habitat increased their absolute areas relative to the first global screening layer. Likely Critical Habitat maintained a largely similar split proportionally across domains but Potential Critical Habitat coverage increased in Areas Beyond National Jurisdiction (ABNJ) relative to land and Exclusive Economic Zones (EEZ). Overall, the updated screening layer increased coverage in all domains, with a large increase from `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="ABNJ","2018_Total"][[1]])` to `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="ABNJ","2025_Total"][[1]])` in ABNJ, `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="EEZ","2018_Total"][[1]])` to `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="EEZ","2025_Total"][[1]])` in EEZ and `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="Land","2018_Total"][[1]])` to `r scales::label_percent(accuracy=0.01)(realm_pcts[realm_pcts$type=="Land","2025_Total"][[1]])` on land.

```{r ch-coverage-table}
#| echo: false
#| tbl-cap: "Areas and percentage coverage of each Critical Habitat category in 2018 and 2025 across realms. Percentages indicate splits within Critical Habitat categories."
#| label: tbl-ch-coverage

wrld_ch_areas = read.csv("wrld_ch_areas.csv",check.names = F) |>
  group_by(type) |>
  summarise(across(starts_with(c("2018","2025")),~sum(.x,na.rm=TRUE)/1000000))

wrld_ch_pct = wrld_ch_areas |>
  reframe(across(2:7,~label_percent(accuracy=.2)(.x/sum(.x)))) |>
  rename_with(~paste0(.x,"_pct"))

wrld_ch_areas_pct = cbind(wrld_ch_areas,wrld_ch_pct) %>% 
  select(type,sort(names(.)))

wrld_ch_areas_pct = cbind(wrld_ch_areas,wrld_ch_pct) %>% 
  select(type,sort(names(.)))

names(wrld_ch_areas_pct) <- NULL

kable(wrld_ch_areas_pct,
      format.args = list(digits=2,big.mark=","),
      format="latex",
      booktabs=T,
      longtable=T) |>
  add_header_above(header=c("Domain"=1,"Likely"=2,"Potential"=2,"Unclassified"=2,
                           "Likely"=2,"Potential"=2,"Unclassified"=2)) |>
  add_header_above(header=c(" "=1,"2018"=6,"2025"=6)) |>
  column_spec(1,width="1cm") |>
  column_spec(c(3,5,7,9,11,13),width="1cm") |>
  column_spec(c(2,4,6,8,10,12),width="0.5cm") |>
  kable_styling(font_size=8) |>
  footnote(general="Areas in km$^2$ x 10$^6$. ABNJ: Areas Beyond National Jurisdiction; EEZ: Economic Exclusion Zone.",escape=F)

```

```{r criteria-areas-table}
#| echo: false
#| tbl-cap: "Areas and percentage cover of each IFC PS6 criteria to Potential and Likely Critical Habitat. Percentages indicate splits within Critical Habitat categories."
#| label: tbl-criteria-areas

criteria_areas = read.csv("drill_down_change_areas.csv", check.names = F) |>
  select(CH,C1:C5,area) |>
  pivot_longer(C1:C5,values_to="present",names_to="criteria") |>
  filter(present>0) |>
  group_by(criteria,CH) |>
  summarise(area=sum(area)/1000,.groups="drop") |>
  pivot_wider(id_cols=1,names_from="CH",values_from="area") |>
  ungroup() |>
  mutate(total_potential=ch2025_areas[ch2025_areas$value==1,2],
         total_likely=ch2025_areas[ch2025_areas$value==10,2],
         likely_pct=label_percent(accuracy=.2)(`10`/(total_likely/1000)),
         potential_pct=label_percent(accuracy=.2)(`1`/(total_potential/1000)),
         criteria=case_match(criteria,
                             "C1" ~ "1. Critically Endangered or Endangered species",
                             "C2" ~ "2. Endemic and/or restricted-range species",
                             "C3" ~ "3. Globally significant concentration of migratory or congregatory species",
                             "C4" ~ "4. Highly threatened and/or unique ecosystems",
                             "C5" ~ "5. Key evolutionary processes")) |>
  select(criteria,`10`,likely_pct,`1`,potential_pct)

kable(criteria_areas,
      col.names = c("Criteria","Likely","","Potential",""),
      format.args = list(digits=2,big.mark=",")) |>
  column_spec(1,width="8cm") |>
  footnote(general="Areas in km$^2$ x 10$^3$. Percentages add up to >100 due to overlapping areas.",escape=F)

```

```{r top-ten}
#| tbl-cap: "States (including associated territories) with the largest areas of identified Critical Habitat."
#| label: tbl-top-ten

wrld_areas = read.csv("O:/f01_projects_active/Global/p08868_CriticalHabitatUpdate/outputs/paper/wrld_ch_areas.csv", check.names=FALSE, encoding="latin1")

wrld_areas |>
    filter(!str_detect(iso_sov,";") & !str_starts(iso_sov,"x")) |>
    group_by(iso_sov) |>
    summarise(likely_total=sum(`2025_Likely`,na.rm=TRUE)/1000,potential_total=sum(`2025_Potential`,na.rm=TRUE)/1000) |>
    mutate(total_ch = likely_total + potential_total) |>
    ungroup() |>
    slice_max(total_ch,n=10) |>
    left_join(dplyr::select(gisco_countrycode,ISO3_CODE,iso.name.en),by=c("iso_sov"="ISO3_CODE")) |>
    mutate(iso.name.en = case_when(iso_sov=="ABNJ" ~ "Areas Beyond National Jurisdiction",.default=iso.name.en)) |>
    dplyr::select(iso.name.en,likely_total,potential_total,total_ch) |>
    kable(col.names = c("State","Likely CH","Potential CH","Total"),
          format.args = list(digits=2,big.mark=",")) |>
    footnote(general="Areas in km$^2$ x 10$^3$.",escape=F)

```

```{r rl-sensitivity}
#| fig-width: 8
#| fig-height: 7
#| fig-cap: Sensitivity analysis of IUCN Red List ranges. Plots show result of removing areas with 3, 10, 20, 50 standard deviations (SDs) higher than the mean of the data. Each plot reports the SD, the number of ranges removed, and the percentage of the total area the data now represent.
#| label: fig-rl-sensitivity
#| cache: false

pal="imola"
revCol=TRUE

rl_full = st_read("rl_full.gpkg",quiet=TRUE) |>
  st_wrap_dateline() |>
  st_transform("EPSG:8857") |>
  st_simplify(dTolerance=10000)

plts = lapply(c(3,10,20,50), function(x){
  plot_df = rl_full %>%
    filter(units::drop_units(st_area(.)) < (x*sd(units::drop_units(st_area(.)))) + mean(units::drop_units(st_area(.))))
  
  area_diff_pct = (sum(units::drop_units(st_area(plot_df)))/sum(units::drop_units(st_area(rl_full))))*100
  
  area_diff_pct = signif(area_diff_pct,3)
  
  row_diff = nrow(rl_full)-nrow(plot_df)
  
  if(revCol){
    rl_pal = rev(scico(palette=pal,n=3))
  } else{
    rl_pal = scico(palette=pal,n=3)
  }
  
  ggplot() +
    geom_sf(data=sphere, fill=NA) +
    geom_sf(data=gisco_coastallines |> st_transform("EPSG:8857"), fill=NA) +
    geom_sf(data=plot_df,aes(fill=category),alpha=0.85,col=NA) +
    scale_fill_manual(values=rl_pal, name=NULL) +
    cowplot::theme_map() +
    labs(title=paste0("SD: ",x),
         subtitle = paste0(area_diff_pct,"% of total area; ",row_diff," ranges missing of ",nrow(rl_full))) +
    theme(legend.position="bottom",
          legend.justification = "centre",
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
})

cowplot::plot_grid(plotlist=plts,nrow=2,ncol=2)

```

```{r raster-sensitivity}
#| fig-width: 5
#| fig-height: 5
#| fig-cap: Sensitivity analysis of % presence cutoff values for mangrove data. Plots show effect of setting the % presence to >25,>50,>75,>90% on the output binary distribution. We select 50% for use in this analysis.
#| label: fig-mangrove-sensitivity
#| cache: false

pal="lipari"
revCol=TRUE

source("plot_functions.R")

plot_rst_sensitivity(data_WGS=rast("../../raw_data/GMW_v3/gmw_v3_2020/mangrove_mosaic_agg.tif"),
             wrld=wrld,
             focus_ext=ext(c(-16.717, -13.634, 10.917, 12.683)),
             palette=pal,
             revCol=revCol)

```

```{r feature-areas-bar}
#| fig-width: 8
#| fig-height: 7
#| fig-cap: Critical Habitat (Likely and Potential) areas per biodiversity feature, split by whether the area is additional, upgraded, unchanged, or downgraded.
#| label: fig-feat-area-bar

feat_bar = read.csv("drill_down_change_areas.csv", check.names=TRUE) |>
  select(c(starts_with(c("L_","P_"))),area,ADDITION,UPGRADE,DOWNGRADE) |>
  pivot_longer(starts_with(c("L_","P_")),names_to="Short",values_to="Presence") |>
  filter(Presence==1) |>
  group_by(Short) |>
  summarise(across(1:4, ~ sum(.x,na.rm=TRUE))) |>
  mutate(NOCHANGE=area-(ADDITION+UPGRADE+DOWNGRADE)) |>
  select(-area) |>
  pivot_longer(ADDITION:NOCHANGE, names_to="Change", values_to="area") |>
  mutate(across(Change, ~ factor(.x,levels=c("DOWNGRADE","NOCHANGE","UPGRADE","ADDITION")))) |>
  left_join(select(read.csv("../../scripts_tools/lookup.csv"),Short,Feature),by="Short") |>
  select(Feature,Change,area)

ggplot(feat_bar, aes(x=fct_reorder(Feature, .x = area, .fun = sum),y=area,fill=Change)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  scale_fill_manual(name=NULL,values=scico(palette="bam",n=5)[2:5],labels=c("Downgraded","No Change","Upgraded","Additional")) +
  coord_flip() +
  ylab(bquote('Area '(km^2))) + xlab(NULL) +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour="gray"),
    panel.grid.minor.x = element_line(colour="gray"),
    axis.text.y  = element_text(hjust=1,size=7),
    panel.background = element_rect(fill="gray80",colour=NA)
    ) +
  guides(fill = guide_legend(reverse=T)) +
  scale_y_continuous(labels=scales::comma,expand=expansion(mult=c(0,.11)))

```

`r ch_changes_areas[ch_changes_areas$ch_type=="ch_Addition",2][[1]]/1000000` million km^2^ has been added, `r ch_changes_areas[ch_changes_areas$ch_type=="ch_Upgrade",2][[1]]/1000000` million km^2^ upgraded, `r ch_changes_areas[ch_changes_areas$ch_type=="ch_Downgrade",2][[1]]/1000000` million km^2^ downgraded, and `r ch_changes_areas[ch_changes_areas$ch_type=="ch_Removal",2][[1]]/1000000` million km^2^ removed in the updated Critical Habitat screening layer.

```{r addition-table}
#| tbl-cap: "Biodiversity features contributing the five highest proportions to the area added as Potential or Likely Critical Habitat in the update."
#| label: tbl-addition

addition_table = read.csv("feat_addition_areas.csv") |>
  rename(Short=Feature) |>
  left_join(read.csv("../../scripts_tools/lookup.csv") |> dplyr::select(Short,Feature),by="Short") |>
  dplyr::select(Feature,2:6)

addition_table |>
  select(-total_add_area) |>
  slice_max(pct_of_add,n=5) |>
  mutate(across(2:3, ~ prettyNum(.x/1000, big.mark=",",digits=4))) |>
  mutate(across(4:5, label_percent(accuracy=0.01))) |>
  select(c(1,3,4,2,5)) |>
  kable(col.names = c("Biodiversity feature",
                      "Area added","%",
                      "Total feature area","%")) |>
  column_spec(1,width="6cm") |>
  column_spec(2:5,width="2cm") |>
  footnote(general="Areas in km$^2$ x 10$^3$. IMMA: Important Marine Mammal Area. Percentages may add up to more than 100% due to overlapping features.", escape=F, threeparttable=TRUE)
  
```

```{r upgrade-table}
#| tbl-cap: "Biodiversity features contributing the five highest proportions to the area upgraded from Potential to Likely Critical Habitat in the update."
#| label: tbl-upgrade

upgrade_table = read.csv("feat_upgrade_areas.csv") |>
  rename(Short=Feature) |>
  left_join(read.csv("../../scripts_tools/lookup.csv") |> dplyr::select(Short,Feature),by="Short") |>
  dplyr::select(Feature,2:6)

upgrade_table |>
  select(-total_upgr_area) |>
  slice_max(pct_of_upgr,n=5) |>
  mutate(across(2:3, ~ prettyNum(.x/1000, big.mark=",",digits=4))) |>
  mutate(across(4:5, label_percent(accuracy=0.01))) |>
  select(c(1,3,4,2,5)) |>
  kable(col.names = c("Biodiversity feature",
                      "Area upgraded","%",
                      "Total feature area","%")) |>
  column_spec(1,width="6cm") |>
  column_spec(2:5,width="2cm") |>
  footnote(general="Areas in km$^2$ x 10$^3$. IBA: Important Bird and Biodiversity Area. Percentages may add up to more than 100% due to overlapping features.", escape=F, threeparttable=TRUE)
  
```

```{r downgrade-table}
#| tbl-cap: "Five largest areas of biodiversity features downgraded from Likely to Potential Critical Habitat in the update. Names do not fully align with features in updated layer."
#| label: tbl-downgrade

downgrade_table = read.csv("feat_2018_downgrade_areas.csv")

downgrade_table |>
  select(-total_down_area) |>
  slice_max(pct_of_down,n=5) |>
  mutate(across(2:3, ~ prettyNum(.x/1000, big.mark=",",digits=4))) |>
  mutate(across(4:5, label_percent(accuracy=0.01))) |>
  select(c(1,3,4,2,5)) |>
  kable(col.names = c("Biodiversity feature",
                      "Area downgraded","%",
                      "Previous feature area","%")) |>
  column_spec(1,width="6cm") |>
  column_spec(2:5,width="2cm") |>
  footnote(general="Areas in km$^2$ x 10$^3$. CR: Critically Endangered; EN: Endangered; KBA: Key Biodiversity Area; IBA: Important Bird and Biodiversity Area. Percentages may add up to more than 100% due to overlapping features.",escape=F,threeparttable=TRUE)
  
```

```{r removal-table}
#| tbl-cap: "Five largest areas of biodiversity features removed as Critical Habitat in the update. Names do not fully align with features in updated layer."
#| label: tbl-removal

removal_table = read.csv("feat_2018_removal_areas.csv")

removal_table |>
  select(-total_remove_area) |>
  slice_max(pct_of_remove,n=5) |>
  mutate(across(2:3, ~ prettyNum(.x/1000, big.mark=",",digits=4))) |>
  mutate(across(4:5, label_percent(accuracy=0.01))) |>
  select(c(1,3,4,2,5)) |>
  kable(col.names = c("Biodiversity feature",
                      "Area removed","%",
                      "Previous feature area","%")) |>
  column_spec(1,width="6cm") |>
  column_spec(2:5,width="2cm") |>
  footnote(general="Areas in km$^2$ x 10$^3$. CR: Critically Endangered; EN: Endangered; PA: protected area; KBA: Key Biodiversity Area. Percentages may add up to more than 100% due to overlapping features.",escape=F,threeparttable=TRUE)
  
```

However, some features used the same data as previously, which warranted further inspection: `r label_percent(accuracy=0.01)(removal_table[removal_table$ALL_FEATURES=="Tropical dry forest","pct_of_feat"][1])` of tropical dry forest has been removed from the layer, alongside `r label_percent(accuracy=0.01)(removal_table[removal_table$ALL_FEATURES=="Soft cold-water coral (Octocorals)","pct_of_feat"][1])` of cold-water coral.

```{r dryfor-sensitivity}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: Differences in tropical dry forest coverage between (a) 2025 aggregation method, (b) original 500m resolution, and (c) 2018 aggregation method.
#| label: fig-dryfor-sensitivity
#| cache: false

cyl = "ESRI:54034"

bol = gisco_get_countries(country="BOL")

dryfor_orig = rast("../../raw_data/WCMC_065_TropicalDryForests2006/Tropical_Dry_Forests/trop_dryf/tropdryf") |>
  crop(bol) |>
  classify(cbind(NA,0))
dryfor2025 = rast("../../scratch/P_C4_Tropical_Dry_Forest.tif") |>
  crop(bol) |>
  disagg(fact=2,method="near") |>
  resample(dryfor_orig,method="near")
dryfor2018 = rast("../../raw_data/WCMC-043-GlobalCH-IFCPS6-2017/01_Data/crhab_mt")|>
  crop(bol |> st_transform(cyl))
crs(dryfor2018) <- cyl

dryfor_vals = cats(dryfor2018)[[1]] |>
  filter(if_any(C1:C5, ~ str_detect(.x, "dry forest"))) |>
  pull(VALUE)

levels(dryfor2018) <- mutate(cats(dryfor2018)[[1]],dryfor=VALUE %in% dryfor_vals)

activeCat(dryfor2018) <- "dryfor"
dryfor2018 = as.numeric(dryfor2018,activeCat(dryfor2018))

dryfor2018 = project(dryfor2018,dryfor2025,method="near") |>
  disagg(fact=2,method="near") |>
  resample(dryfor_orig,method="near") |>
  classify(cbind(NA,0))

stack = c(dryfor2025,dryfor_orig,dryfor2018)
names(stack) <- letters[1:3]
levels(stack) <- list(data.frame(id=0:1,a=c("No","Yes")),
                      data.frame(id=0:1,b=c("No","Yes")),
                      data.frame(id=0:1,c=c("No","Yes")))

focus_ext = ext(bol)
inset_ext = focus_ext+25

a = ggplot() +
  geom_spatraster(data = stack) +
  facet_wrap(~lyr, ncol=3) +
  theme_void() +
  theme(legend.position="bottom") +
  scale_fill_manual(name="Tropical dry forest",values=scico(palette=pal,direction=ifelse(revCol,-1,1),n=3)[2:3])

b = ggplot(st_transform(wrld,4326)) +
  geom_sf(col=NA) +
  geom_sf(data=st_as_sfc(st_bbox(focus_ext,crs=4326)),col="red",fill=NA) +
  theme_map() +
  theme(panel.background = element_rect(fill="gray",colour=NA)) +
  coord_sf(xlim = inset_ext[1:2], ylim = inset_ext[3:4], expand = FALSE)

ggdraw(a) +
  draw_plot(b,scale=0.265,halign=0,valign=0)

```

```{r octo-sensitivity}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: Differences in octocoral coverage between (a) 2025 threshold (>90%), (b) >=90% threshold, and (c) 2018 octocoral coverage.
#| label: fig-octo-sensitivity
#| cache: false

espmar = ext(c(-4,-2,35,36.5))

coral_files = list.files("../../raw_data/ZSL-001-ModelledOctocorals2012/02_Data_sources/Restricted-GeoTiffHighRes", full.names=TRUE) %>% 
  keep(str_ends(.,".tif") & !str_detect(.,"Consensus"))

octo2018 = rast("ch2018_drill_down.tif") |>
  crop(espmar)

octo_vals = cats(octo2018)[[1]] |>
  filter(if_any(C1:C5, ~ str_detect(.x, "Octocorals"))) |>
  pull(VALUE)

levels(octo2018) <- mutate(cats(octo2018)[[1]],octo=VALUE %in% octo_vals)

activeCat(octo2018) <- "octo"
octo2018 = as.numeric(octo2018,activeCat(octo2018))

coral = rast(coral_files) |>
  crop(espmar) |>
  classify(cbind(NA,0))

stack = c(app(coral>90,any),app(coral>=90,any),octo2018)
names(stack) <- letters[1:3]
levels(stack) <- list(data.frame(id=0:1,a=c("No","Yes")),
                      data.frame(id=0:1,b=c("No","Yes")),
                      data.frame(id=0:1,c=c("No","Yes")))

focus_ext = ext(espmar)
inset_ext = focus_ext+25

a = ggplot() +
  geom_spatraster(data = stack) +
  geom_sf(data=st_crop(wrld,st_bbox(extend(focus_ext,y=0.5))),col=scico(palette=pal,n=3)[2],fill=NA) +
  facet_wrap(~lyr, ncol=3) +
  theme_void() +
  theme(legend.position="bottom") +
  scale_fill_manual(name="Octocorals",values=scico(palette=pal,direction=ifelse(revCol,-1,1),n=3)[2:3]) +
  coord_sf(xlim = focus_ext[1:2], ylim = focus_ext[3:4], expand = FALSE)

b = ggplot(st_transform(wrld,4326)) +
  geom_sf(col=NA) +
  geom_sf(data=st_as_sfc(st_bbox(focus_ext,crs=4326)),col="red",fill=NA) +
  theme_map() +
  theme(panel.background = element_rect(fill="gray",colour=NA)) +
  coord_sf(xlim = inset_ext[1:2], ylim = inset_ext[3:4], expand = FALSE)

ggdraw(a) +
  draw_plot(b,scale=0.28,halign=0,valign=0)

```